Live Functional Programming with Typed Holes

### 2018-02-20 examples.tex
Section 2, wip

@@

\overviewExample{5}{\rkc{...}}

\rkc{gradual typing error, cast error, run around}

\begin{lstlisting}
foo : bool -> ? -> int
foo b =
  if b then
    x + 10
  else
    x 10

foo false 1 + 2 + 2
\end{lstlisting}

this demonstrates how, when no expression holes, our system is like gradual
typing, but with the ability to run around cast errors.
%
just as it may be useful to report multiple type errors, this allows ability to
report multiple dynamic errors, as well additional dynamic behavior around them.
%
could be incorporated into other gradual type systems

>>>>

An alternative to marking type-inconsistent expressions as sources of error
(with non-empty holes) is to use \emph{gradual types}~(\eg{}~\cite{XXX,XXX,XXX})
to sidestep the choice of a assigning single static type, instead deferring to
dynamic casts.
%
\HazelnutLive{} inherits the notion of \emph{type holes} from
\citet{popl-paper}, which serves a similar static purpose as the unknown type in
gradual type systems.

\overviewExample{5}{Dynamically Typed Negation}

Consider the \li{negate} function (adapted from \cite{ChughPOPL2012}), which
does not have a conventional static type (without using expression holes),
because the argument \li{x} is used at type \li{int} on line \rkc{XXX} and
function type on line \rkc{XXX}.
%
Therefore, the declared type of \li{x} is the hole type \li{??}, which allows
\li{x} to be used at conflicting types by expanding each use with \emph{casts}
that ensure the safety of each dynamic use.
%
The return type also uses the hole type \li{??}.

\begin{lstlisting}
negate : bool -> ?? -> ??
negate b x =
  if b
    then 0 - x     // expanded to: (0 - (x<?? => int>))<int => ??>
    else not x     // expanded to: (not (x<?? => bool>))<bool => ??>

(negate false 1) + 2 + 3 + (negate true 4)
\end{lstlisting}

\noindent
%
As in prior gradually typed languages~\cite{XXX},
evaluating the expression \li{negate true 1} on line \rkc{XXX} leads to
\li{(not (1<int => ??><?? => bool>))<bool => ??>}, where the inner casts
lead to the \emph{cast error} \li{1<?? =/=> bool>} because \li{1} cannot be
safely treated as having type \li{bool}.
%
Unlike in prior gradually typed languages, however, \HazelnutLive{} evaluates
around the cast error, making progress on the \li{2 + 3 + negate false 4}
expression that surrounds the error; the final indeterminate result is
\li{(not (1<?? =/=> bool>))<bool => ??> + 1}.
%
Just like it is useful for static type checkers to report multiple errors, our
approach allows us to report multiple dynamic cast errors, and otherwise make
progress on expressions that do not depend on failed casts.
%
This approach can be incorporated into existing gradually typed languages.

@@
